---
title: "Analysis of Shark Attacks in the USA from 1960 to 2018."
author: "Yaakov Haiby"
date: "2023-08-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r, echo = FALSE, results = 'hide', warning=FALSE, message=FALSE}

library(tidyverse)
library(ggplot2)
library(maps)
library(mapproj)
library(rworldmap)
library(RColorBrewer)




```
<br>**A link to the data source :** "https://www.kaggle.com/datasets/teajay/global-shark-attacks"


### Shark Attack Frequencies Over the Years.

```{r}
path <- "/Users/yaakovhaiby/Downloads/attacks.csv"
data <- read.csv(path)
data <- filter(data, Country == "USA")
data <- filter(data,Year>=1960)
```


<br>**code explanation :**

We delete the cases where the attack was questionable from the data, as we are interested in only those cases where the attack was definitively caused by a shark.
```{r}
# Define the strings to filter
strings_to_remove <- c("Questionable incident", "Shark involvement not confirmed", "No shark involvement","No shark invovlement - it ws a publicity stunt")

# Filter out rows where the "Species" column contains any of the specified strings
data <- data %>% filter(!(Species %in% strings_to_remove))

```


```{r, warning=FALSE, message=FALSE}

years_vec <- unlist(data$Year)
years_table <- table(years_vec)
years_table <- as.data.frame(years_table)
years_table <- years_table %>% arrange(desc(years_table)) 


# Convert the table to a data frame and rename the variable to Var1
names(years_table)[1] <- "Var1"
years_table$Var1 <- as.numeric(as.character(years_table$Var1))

# Sort the data frame in ascending order based on the years
years_table <- years_table %>% arrange(Var1)

ggplot(years_table, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity", fill = "black", color = "white") +
  geom_smooth(method = "loess", se = FALSE, color = "red", size = 1.5) +
  labs(title = "Frequency of Years",
       x = "Year",
       y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(limits = c(1960, 2018), breaks = seq(1960, 2018, 10))

```

### Monthly Cumulative Case Counts Throughout the Year.

```{r}
# Function to keep only English letters in a string

data1 <- data

keep_english_letters <- function(text) {
  gsub("[^A-Za-z]", "", text)
}

# Apply the function to the 'Date' column
data1$Date <- lapply(data1$Date, keep_english_letters)

# View the updated data1 frame

# Assuming you have already read the CSV file into the 'data1' data1 frame

# Filter out rows where the "Date" column has more than 3 letters
data1 <- data1[nchar(data1$Date) <= 3, ]

# View the updated data1 frame
data1 <- data1 %>% arrange(Date)

```


```{r}

# Assuming you have already read the CSV file into the 'data1' data1 frame

# Create a vector with the three-letter shortcuts for months
months_short <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Filter out rows where the "Date" column contains at least one of the elements in months_short
data1 <- data1 %>% 
  filter(!is.na(Date) & !Date %in% c("", "NA") & any(grepl(paste(months_short, collapse = "|"), Date)))

```


```{r}
# Create a vector with the three-letter shortcuts for months
months_short <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Filter out rows where the "Date" column does not contain any of the elements in months_short
data1 <- data1 %>%
  filter(grepl(paste(months_short, collapse = "|"), Date))

# View the updated data1 frame

data1 <- data1 %>% arrange(Date,Area)
Months <-  unlist(data1$Date)

element_counts <- table(Months)
element_counts <- as.data.frame(element_counts)
# Define the correct order of months
month_order <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Convert the "Month" column to an ordered factor with specified levels
element_counts$Month <- factor(element_counts$Month, levels = month_order, ordered = TRUE)


# Create the plot using ggplot2
ggplot(element_counts, aes(x = Month, y = Freq)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Number of comulative cases in each month 1960-2018 ",
       x = "Month",
       y = "Number of Cases") +
  theme_minimal()
```

### Visualizing Cumulative Shark Attack Cases on a Map of the USA.


```{r}
data <- data %>% arrange(Area)


States_vec <- unlist(data$Area)
invisible(States_data <- table(States_vec))
States_data <- as.data.frame(States_data)
names(States_data)[2] <- "cumulative_attacks"
States_data <- States_data %>% arrange(desc(States_data$`Cumulative attacks`))
names(States_data)[1]<- "region"
States_data$region <- as.character(States_data$region)

# Assuming your data is named States_data and is already loaded

# Convert any empty region names to NA
States_data$region[States_data$region == ""] <- NA

# Remove leading/trailing whitespace from region names
States_data$region <- trimws(States_data$region)

summarized_data <- States_data %>%
  group_by(region) %>%
  summarise(total_cumulative_attacks = sum(cumulative_attacks, na.rm = TRUE))
summarized_data <- summarized_data %>%
  filter(!is.na(region) & !is.na(total_cumulative_attacks))

States_data <-summarized_data



states <- map_data("state")
attacks <- States_data

# Assuming "cumulative_attacks" is the column with attack data
names(attacks) <- tolower(names(attacks))
attacks$region <- tolower(attacks$region)
# Find missing countries
attacks.geo <- merge(states, attacks, sort = FALSE, by = "region")
attacks.geo <- attacks.geo[order(attacks.geo$order), ]

# Find missing countries
missing_countries <- setdiff(tolower(states$region), tolower(attacks$region))

# Create a plot of missing countries in white
plot_missing <- ggplot() +
  geom_polygon(data = states[states$region %in% missing_countries, ],
               aes(x = long, y = lat, group = group),
               fill = "white", color = "gray") +
  theme_void()

# Combine the missing countries plot with attacks.geo
final_plot <- plot_missing +
  geom_polygon(data = attacks.geo,
               aes(x = long, y = lat, group = group, fill = total_cumulative_attacks)) +
  scale_fill_gradient(low = "lightgreen", high = "red", na.value = "white") +
  coord_map()

print(final_plot)






```

### Pie Chart: Proportion of Cumulative Shark Attacks by Species Among the Top 7 Most Attacked Shark Species.

```{r,warning=FALSE}
data2<- data

data2$Species[grepl("white", data2$Species, ignore.case = TRUE)] <- "Great white shark"
data2$Species[grepl("Tiger", data2$Species, ignore.case = TRUE)] <- "Tiger shark"
data2$Species[grepl("Bull", data2$Species, ignore.case = TRUE)] <- "Bull shark"
data2$Species[grepl("Nurse", data2$Species, ignore.case = TRUE)] <- "Nurse shark"
data2$Species[grepl("Blacktip", data2$Species, ignore.case = TRUE)] <- "Blacktip shark"
data2$Species[grepl("Mako", data2$Species, ignore.case = TRUE)] <- "Mako shark"
data2$Species[grepl("hammerhead", data2$Species, ignore.case = TRUE)] <- "Hammerhead shark"
data2$Species[grepl("Spinner", data2$Species, ignore.case = TRUE)] <- "Spinner shark"
data2$Species[grepl("Leopard", data2$Species, ignore.case = TRUE)] <- "Leopard shark"
data2$Species[grepl("Sandbar", data2$Species, ignore.case = TRUE)] <- "Sandbar shark"
data2$Species[grepl("Sevengill", data2$Species, ignore.case = TRUE)] <- "Sevengill shark"
data2$Species[grepl("Shovelnose", data2$Species, ignore.case = TRUE)] <- "Shovelnose shark"
data2$Species[grepl("sand shark", data2$Species, ignore.case = TRUE)] <- "sand shark"
data2$Species[grepl("Lemon", data2$Species, ignore.case = TRUE)] <- "Lemon shark"
data2$Species[grepl("Blue", data2$Species, ignore.case = TRUE)] <- "Blue shark"
data2$Species[grepl("Silky", data2$Species, ignore.case = TRUE)] <- "Silky shark"
data2$Species[grepl("Reef", data2$Species, ignore.case = TRUE)] <- "Reef shark"
data2$Species[grepl("Salmon", data2$Species, ignore.case = TRUE)] <- "Salmon shark"

```



```{r}
shark_species <- c("Great white shark", "Tiger shark", "Bull shark", "Nurse shark",
                   "Blacktip shark", "Mako shark", "Hammerhead shark", "Spinner shark",
                   "Leopard shark", "Sandbar shark", "Sevengill shark", "Shovelnose shark",
                   "sand shark", "Lemon shark", "Blue shark", "Silky shark", "Reef shark","Salmon shark")

```


```{r}
# Convert the data2$Species vector to lowercase for case-insensitive comparison
data2$Species <- iconv(na.omit(data2$Species), to = "ASCII//TRANSLIT")
data2$Species <- tolower(data2$Species)
# Function to check if any substring from shark_species is present in the input string
check_substring <- function(x) {
  any(grepl(paste(shark_species, collapse = "|"), x, ignore.case = TRUE))
}

# Apply the check_substring function to each row of the "Species" column
keep_rows <- apply(data2, 1, function(row) check_substring(row["Species"]))

# Subset the data frame to keep only the rows with at least one valid substring
data2 <- data2[keep_rows, ]

# If needed, convert the data2$Species back to title case (optional)
data2$Species <- tools::toTitleCase(data2$Species)

```


```{r}
shark_vec <- unlist(data2$Species)
invisible(shark_table <- table(shark_vec))
shark_table <- as.data.frame(shark_table)
shark_table <- shark_table %>% arrange(desc(Freq)) 


```



```{r}
# Create a copy of the shark_table data frame
shark_table_modified <- shark_table

# Calculate the percentages
shark_table_modified$percentage <- (shark_table_modified$Freq / sum(shark_table_modified$Freq)) * 100

# Sort the data frame by percentage in descending order
shark_table_modified <- shark_table_modified[order(-shark_table_modified$percentage), ]

# Create a pie chart with different colors for each species
p <- ggplot(shark_table_modified, aes(x = "", y = Freq, fill = shark_vec)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  scale_fill_discrete() +  # Use different colors for each species
  labs(title = "Proportion of Cumulative Attacks from All Shark Species",
       x = NULL,
       y = NULL,
       fill = "Shark Species") +
  theme_void() +
  theme(legend.position = "right")

# Calculate the threshold for the top 7
threshold <- shark_table_modified$percentage[7]

# Add percentages as labels only for the top 7 species, and combine the rest as "Other"
shark_table_modified$shark_vec <- ifelse(shark_table_modified$percentage >= threshold,
                                         as.character(shark_table_modified$shark_vec),
                                         "Other")

# Recreate the pie chart with modified labels
p <- ggplot(shark_table_modified, aes(x = "", y = Freq, fill = shark_vec)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  scale_fill_discrete() +
  labs(title = "Proportion of Cumulative Attacks from top 7 Most Attacked Shark Species",
       x = NULL,
       y = NULL,
       fill = "Shark Species") +
  geom_text(aes(label = ifelse(shark_vec == "Other", "", paste(round(percentage, 1), "%"))),
            position = position_stack(vjust = 0.5)) +
  theme_void() +
  theme(legend.position = "right")

# Print the updated pie chart
print(p)




```

### Highlighting the Predominant Shark Species Responsible for Attacks in Each Region.






```{r}

# Assuming you have a data frame named data2 with columns "Area" and "Species"

# Calculate the most frequently occurring attacking shark species for each area
most_common_attacker <- data2 %>%
  group_by(Area, Species) %>%
  summarize(Frequency = n(), .groups = 'drop') %>%
  arrange(Area, desc(Frequency)) %>%
  group_by(Area) %>%
  slice(1) %>%
  ungroup()

# Custom color palette with more colors
species_colors <- c("#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3", "#a6d854",
                    "#ffd92f", "#e5c494", "#b3b3b3", "#1f78b4", "#33a02c",
                    "#984ea3", "#ff7f00", "#cab2d6", "#6a3d9a", "#fb9a99",
                    "#fdbf6f", "#b15928", "#ffff99")

# Create a bar plot showing the most common attacking shark species for each area
ggplot(data = most_common_attacker, aes(x = Area, y = Frequency, fill = Species)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = species_colors) +
  labs(title = "Most Common Attacking Shark Species in Each Area",
       x = "Area",
       y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```



### Displaying the Incidence of Shark Species Attacks in Florida.


```{r}
#who is the shark that attacks the most in Florida?
library(ggplot2)
library(dplyr)

# Assuming you have a data frame named data2 with columns "Area" and "Species"

data3 <- data2 %>% filter(Area == "Florida")
shark_vec2 <- unlist(data3$Species)
shark_table2 <- table(shark_vec2)
shark_table2 <- as.data.frame(shark_table2)
shark_table2 <- shark_table2 %>%
  arrange(desc(Freq))

# Create a more visually appealing bar plot with a different color palette
ggplot(shark_table2, aes(x = reorder(shark_vec2, -Freq), y = Freq, fill = shark_vec2)) +
  geom_bar(stat = "identity") +
  labs(title = "Frequency of Shark Species in Florida",
       x = "Shark Species",
       y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +  # Hide the legend for individual bars
  scale_fill_viridis_d(option = "D", begin = 0.3, end = 0.8)  # Use the Category20 color palette

```

### Displaying the Incidence of Shark Species Attacks in California.



```{r}
#who is the shark that attacks the most in California ?
library(ggplot2)
library(dplyr)

# Assuming you have a data frame named data2 with columns "Area" and "Species"

data4 <- data2 %>% filter(Area == "California")
shark_vec3 <- unlist(data4$Species)
shark_table3 <- table(shark_vec3)
shark_table3 <- as.data.frame(shark_table3)
shark_table3 <- shark_table3 %>%
  arrange(desc(Freq))

# Create a horizontal bar plot
ggplot(shark_table3, aes(x = Freq, y = reorder(shark_vec3, Freq), fill = shark_vec3)) +
  geom_bar(stat = "identity") +
  labs(title = "Frequency of Shark Species in California",
       x = "Frequency",
       y = "Shark Species") +
  theme_minimal() +
  theme(axis.text.y = element_text(hjust = 0)) +
  scale_fill_brewer(palette = "Set3")


```

### Analyzing the Frequency of Shark Attacks During Various Aquatic Recreational Activities.


```{r,warning=FALSE,}

data5 <-data

data5 <- subset(data5, !is.na(Activity) & Activity != "")


data5$Activity[grepl("Swimming", data5$Activity, ignore.case = TRUE)] <- "Swimming"
data5$Activity[grepl("Kitesurfing", data5$Activity, ignore.case = TRUE)] <- "Surfing"
data5$Activity[grepl("Boating", data5$Activity, ignore.case = TRUE)] <- "Boating"
data5$Activity[grepl("Hunting", data5$Activity, ignore.case = TRUE)] <- "Hunting or fishing"
data5$Activity[grepl("Fishing", data5$Activity, ignore.case = TRUE)] <- "Hunting or fishing"
data5$Activity[grepl("Surfing", data5$Activity, ignore.case = TRUE)] <- "Surfing"
data5$Activity[grepl("diving", data5$Activity, ignore.case = TRUE)] <- "Scuba diving"
data5$Activity[grepl("Body boarding", data5$Activity, ignore.case = TRUE)] <- "Surfing"
data5$Activity[grepl("Playing", data5$Activity, ignore.case = TRUE)] <- "Bathing"
data5$Activity[grepl("bathing", data5$Activity, ignore.case = TRUE)] <- "Bathing"
data5$Activity[grepl("Snorkeling", data5$Activity, ignore.case = TRUE)] <- "Scuba diving"
data5$Activity[grepl("Wading", data5$Activity, ignore.case = TRUE)] <- "bathing"
data5$Activity[grepl("Killing", data5$Activity, ignore.case = TRUE)] <- "Hunting or fishing"
data5$Activity[grepl("tagging", data5$Activity, ignore.case = TRUE)] <- "An attack during research"
data5$Activity[grepl("Shooting", data5$Activity, ignore.case = TRUE)] <- "Hunting or fishing"
data5$Activity[grepl("paddling", data5$Activity, ignore.case = TRUE)] <- "Surfing"
data5$Activity[grepl("Walking", data5$Activity, ignore.case = TRUE)] <- "Bathing"
data5$Activity[grepl("Boat", data5$Activity, ignore.case = TRUE)] <- "Boating"

vec <- c("Swimming","Surfing","Boating","Hunting or fishing","Scuba diving","bathing","An attack during research")
data5$Activity <- ifelse(data5$Activity %in% vec, data5$Activity, "Other")

unique_activities <- unique(data5$Activity)


activity_vec <- unlist(data5$Activity)
activity_table <- table(activity_vec)
activity_table <- as.data.frame(activity_table)
activity_table <- activity_table %>% arrange(desc(Freq))

```




```{r}
# Assuming you have your own activity_table with columns "activity_vec" and "Freq"

# Calculate percentages and cumulative percentages
activity_table$fraction = activity_table$Freq / sum(activity_table$Freq)
activity_table$ymax = cumsum(activity_table$fraction)
activity_table$ymin = c(0, head(activity_table$ymax, n = -1))

# Create the donut plot
donut_plot <- ggplot(activity_table, aes(ymax = ymax, ymin = ymin, xmax = 4, xmin = 3, fill = activity_vec)) +
  geom_rect() +
  coord_polar(theta = "y") +
  xlim(c(2, 4)) +
  labs(title = "frequency of Shark Attacks during the activity ",
       fill = "Activity",
       caption = paste("Total:", sum(activity_table$Freq))) +
  theme_void() +
  theme(legend.position = "right")

# Show the donut plot
print(donut_plot)

```

### Illustrating the Frequency of Shark Species Attacks During Various Human Activities.





```{r,warning=FALSE}
data7 <- data2
data7 <- subset(data7, !is.na(Activity) & Activity != "")


data7$Activity[grepl("Swimming", data7$Activity, ignore.case = TRUE)] <- "Swimming"
data7$Activity[grepl("Kitesurfing", data7$Activity, ignore.case = TRUE)] <- "Surfing"
data7$Activity[grepl("Boating", data7$Activity, ignore.case = TRUE)] <- "Boating"
data7$Activity[grepl("Hunting", data7$Activity, ignore.case = TRUE)] <- "Hunting or fishing"
data7$Activity[grepl("Fishing", data7$Activity, ignore.case = TRUE)] <- "Hunting or fishing"
data7$Activity[grepl("Surfing", data7$Activity, ignore.case = TRUE)] <- "Surfing"
data7$Activity[grepl("diving", data7$Activity, ignore.case = TRUE)] <- "Scuba diving"
data7$Activity[grepl("Body boarding", data7$Activity, ignore.case = TRUE)] <- "Surfing"
data7$Activity[grepl("Playing", data7$Activity, ignore.case = TRUE)] <- "Bathing"
data7$Activity[grepl("bathing", data7$Activity, ignore.case = TRUE)] <- "Bathing"
data7$Activity[grepl("Snorkeling", data7$Activity, ignore.case = TRUE)] <- "Scuba diving"
data7$Activity[grepl("Wading", data7$Activity, ignore.case = TRUE)] <- "bathing"
data7$Activity[grepl("Killing", data7$Activity, ignore.case = TRUE)] <- "Hunting or fishing"
data7$Activity[grepl("tagging", data7$Activity, ignore.case = TRUE)] <- "An attack during research"
data7$Activity[grepl("Shooting", data7$Activity, ignore.case = TRUE)] <- "Hunting or fishing"
data7$Activity[grepl("paddling", data7$Activity, ignore.case = TRUE)] <- "Surfing"
data7$Activity[grepl("Walking", data7$Activity, ignore.case = TRUE)] <- "Bathing"
data7$Activity[grepl("Boat", data7$Activity, ignore.case = TRUE)] <- "Boating"

vec <- c("Swimming","Surfing","Boating","Hunting or fishing","Scuba diving","bathing","An attack during research")
data7$Activity <- ifelse(data7$Activity %in% vec, data7$Activity, "Other")

```



```{r,warning=FALSE}

# Assuming you have a data frame named data7 with columns "Activity" and "Species"

# Calculate the frequency of each shark species for each activity
species_activity_freq <- data7 %>%
  group_by(Activity, Species) %>%
  summarize(Frequency = n(), .groups = 'drop') %>%
  ungroup()

# Create a color palette for distinct species
color_palette <- scales::hue_pal()(length(unique(species_activity_freq$Species)))

# Create a bar plot showing the frequency of shark species for each activity
ggplot(data = species_activity_freq, aes(x = Activity, y = Frequency, fill = Species)) +
  geom_bar(stat = "identity") +
  labs(title = "Shark Species Frequency for Each Activity",
       x = "Activity",
       y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = color_palette)


```



### Attacks Frequency by Daytime.


```{r,warning=FALSE,message=FALSE,error=FALSE}
# Assuming your data frame is named 'data'
shark_attacks_df <- subset(data, !is.na(Time))

shark_attacks_df <- shark_attacks_df %>%
  mutate(Time_Category = case_when(
    is.na(Time) | Time == "" | !grepl("^\\d\\d[hH]\\d\\d$", Time) ~ Time,
    between(as.numeric(substr(Time, 1, 2)), 5, 11) ~ "morning",
    between(as.numeric(substr(Time, 1, 2)), 12, 15) ~ "noon",
    between(as.numeric(substr(Time, 1, 2)), 16, 18) ~ "afternoon",
    between(as.numeric(substr(Time, 1, 2)), 19, 23) ~ "evening",
    TRUE ~ "night"
  ))


# Modify specific values in the Time_Category column
shark_attacks_df <- shark_attacks_df %>%
  mutate(Time_Category = case_when(
    Time_Category %in% c(
      "10h45-11h15", "07h00 - 08h00", "Morning", "Early Morning",
      "10h00 / 11h00", "08h00 / 09h30", "Morning ", "Just before noon",
      "11h115", "Early morning", "Sunset", "8:04 pm", "Mid morning",
      "Prior to 10h37", "Daybreak", "Mid-morning", "\"Just before 11h00\"",
      ">06h45", "<07h30", "09h30 ", "11h30 ", "Shortly before 13h00",
      "09h30 / 10h00", "Between 05h00 and 08h00", "   ",
      " 14h00", "\"After lunch\"", "12h45 / 13h45", "1300", "15j45",
      "Noon", "14h00  -15h00", "13h345", "14h00 - 15h00", "15h00j",
      "14h30 / 15h30"
    ) ~ "morning",
    Time_Category %in% c(" 14h00", "\"After lunch\"", "12h45 / 13h45",
                         "1300", "15j45", "Noon", "14h00  -15h00",
                         "13h345", "14h00 - 15h00", "15h00j") ~ "noon",
    Time_Category %in% c("After noon", "Early afternoon", "1600",
                         "18h30?", "Late afternon", "Late Afternoon",
                         "18h15-18h30", "Afternoon") ~ "afternoon",
    Time_Category %in% c("Evening", "Dusk", "Late afternoon",
                         "Just before sundown") ~ "evening",
    Time_Category %in% c("Night", "03h45 - 04h00", "Shortly after midnight",
                         "\"Night\"", "Midnight") ~ "night",
    TRUE ~ Time_Category
  )) %>%
  filter(!Time_Category %in% c("09h30 / 15h30", "10h30 or 13h30", 
                               "10h00 or 14h00", " ", "--", "P.M.", "X", "A.M.",
                               "Possibly same incident as 2000.08.21", ""))

day_vec <- unlist(shark_attacks_df$Time_Category)
invisible(day_table <- table(day_vec))
day_table <- as.data.frame(day_table)
day_table <- day_table %>% arrange(desc(Freq))
```


```{r,warning=FALSE,message=FALSE,error=FALSE}

# Define the desired order of 'day_vec' levels
desired_order <- c("morning", "noon", "afternoon", "evening", "night")

# Reorder the levels of 'day_vec' factor
day_table$day_vec <- factor(day_table$day_vec, levels = desired_order)

# Define your custom color palette
color_palette <- c("morning" = "blue", "noon" = "green", "afternoon" = "orange", "evening" = "purple", "night" = "black")

# Create a lollipop plot using ggplot
ggplot(day_table, aes(x = day_vec, y = Freq)) +
  geom_segment(aes(xend = day_vec, yend = 0), color = "gray") +
  geom_point(aes(color = day_vec), size = 3, position = position_dodge(width = 0.5)) +
  scale_color_manual(values = color_palette) +
  labs(title = "Attacks Frequency by Daytime",
       x = "Daytime",
       y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```



### Fatal shark attacks:

#### Based on the data, what is the likelihood of a fatal shark attack?


```{r}
# Assuming your data frame is named 'shark_attacks_df'
# Replace this with the actual name of your data frame

# Step 1: Count the number of fatal attacks
fatal_attacks_count <- sum(data$Fatal..Y.N. == "Y")

# Step 2: Count the total number of attacks
total_attacks_count <- nrow(data)

# Step 3: Calculate the empirical probability of getting killed
empirical_probability <- fatal_attacks_count / total_attacks_count

empirical_probability <- round(empirical_probability,3)

print(paste("Empirical Probability of Getting Killed by a Shark:", empirical_probability))

```


#### Empirical Probability of Fatal Shark Attacks by Species.


```{r}
# Assuming your data frame is named data2

# Calculate empirical probabilities for all shark species
species_probabilities <- data2 %>%
  group_by(Species) %>%
  summarize(empirical_probability = sum(Fatal..Y.N. == "Y") / n())

# Create a bar plot
ggplot(species_probabilities, aes(x = reorder(Species, -empirical_probability), y = empirical_probability)) +
  geom_bar(stat = "identity", fill = "dodgerblue") +
  theme_minimal() +
  coord_flip() +
  labs(x = "Shark Species", y = "Empirical Probability of Fatal Attacks",
       title = "Empirical Probability of Fatal Attacks by Shark Species")

```



#### Fatalities from Shark Attacks Over the Years.





```{r}
data9<-data
filtered_data <- data9[data9$Fatal..Y.N. == "Y", ]
filtered_data <- filtered_data[filtered_data$Year >= 1960 & filtered_data$Fatal..Y.N. == "Y",]
year_counts <- filtered_data %>%
  group_by(Year) %>%
  summarize(Count = n())
# Create a bar plot with a built-in color scale
ggplot(year_counts, aes(x = factor(Year), y = Count, fill = factor(Year))) +
  geom_bar(stat = "identity") +
  labs(title = "Occurrences of Each Year",
       x = "Year",
       y = "Count") +
  scale_fill_viridis_d() +  # Use a built-in color scale
  theme_minimal() +
  scale_x_discrete(breaks = unique(year_counts$Year[seq(1, nrow(year_counts), by = 2)]))

```







